Variational Monte Carlo on Bosonic Systems

### 2018-04-03 project_1/article/article.tex
Is it fixed??
Summary remarks, abstract.

\title{Variational Monte Carlo on bosonic systems}
\author{Winther-Larsen, Sebastian Gregorius}
\homepage[Project code: ]{https://github.com/Schoyen/FYS4411}
\affiliation{University of Oslo}
\author{Schøyen, Øyvind Sigmundson}
\homepage[Project code: ]{https://github.com/Schoyen/FYS4411}
\affiliation{University of Oslo}
\date{\today}

\begin{abstract}
    In this project we use Quantum Variational Monte Carlo on bosonic systems.
    \begin{figure}[h!]
        \centering
        \includegraphics[width=0.7\textwidth]{loopholes.jpg}
    \end{figure}
\end{abstract}

@@

        The Laplacian of particle $k$ divided by the trial wavefunction is
        \begin{widetext}
            \begin{align}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                &=
                \frac{\nabla_k^2\phi_k}{\phi_k}
                + 2\frac{\nabla_k\phi_k}{\phi_k}
                \sum_{m \neq k}^N
                \frac{\Delta\vf{r}_{km}}{r_{km}}
                \dpd[]{u_{km}}{r_{km}}
                + \sum_{m\neq k}^N
                \para{
                    \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                    + \dpd[2]{u_{km}}{r_{km}}
                }
                +
                \para{
                    \sum_{m \neq k}^N
                    \frac{\Delta\vf{r}_{km}}{r_{km}}
                    \dpd[]{u_{km}}{r_{km}}
                }^2,
                \label{eq:laplacian_full_system}
            \end{align}
        \end{widetext}
        where $\Delta \vf{r}_{km} = \vf{r}_k - \vf{r}_m$ and $d$ is the
        dimensionality. Evaluating the gradient and Laplacian of the single
        particle functions yields
        \begin{gather}

            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)
        \Psi_T(\vf{r}).
        \label{eq:gradient_full_wavefunction}
    \end{align}

@@

        Looking at the one-body part of the Hamiltonian we get
        \begin{align}
            H
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}\nabla_k^2
                + \half m\brak{
                    \omega^2(x_k^2 + y_k^2)
                    + \omega_z^2 z_k^2
                }
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
                },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The single particle functions also get a scaling
            factor
            \begin{align}
                \phi_k
                &=
                \exp\brak{
                    -\frac{\alpha\hbar}{m\omega}
                    ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
                }
                \equiv
                \phi(\vf{r}_k').
                \label{eq:scaled_spf}
            \end{align}
            The correlation wavefunction and the interaction potential remain the
            same, but using the scaled hard core radius $a/a_{\text{ho}} =
            0.0043$.

@@

        \subsubsection{One-Body Densities}
            We now look at one-body densities for the value of $\alpha$ found in
            \autoref{eq:optimal_interacting_alpha}. The results for a system consisting
            of $n=100$ particles are displayed in \autoref{fig:one_body_density_100}.
            In this figure we see the densities for for a system confined to a 
            perturbed elliptic trap with and without interaction. We see that for the
            interacting case, the one-body density is skewed towards a larger 
            $\abs{\vf{r}}$, as is expected. The `ìdeal'' density, a system with spherical
            trap, is also included in the figure.

@@

    \subsection{Energy per particle}
        For the non-interacting harmonic oscillator system there is no
        difference between a new ``particle'' and new dimensions
        \footnote{This is not entirely true in the case of our implementation of
        the Metropolis algorithms as we draw a single random particle and move
        all dimensions instead of iterating over all particles and all
        dimensions.}. This means that the inclusion of a new particle or
        dimension yields a linear scaling in the energy. Thus if we compute the
        energy per particle for systems of varying size we expect to get the
        same energy. This is most easily seen in \autoref{fig:initial_problem_b}
        and in the expression for the exact energy \autoref{eq:exact_energy}.
        This, however, is \emph{not} the case in the interacting system due to
        the Jastrow factor. Looking at \autoref{tab:10_interacting} and
        \autoref{tab:100_interacting} we see that multiplying the expected
        energy of theformer with $10$ yields a lower number than the latter. We
        can also see this behavior in \autoref{fig:interacting}, where the
        energy per particle for $N = \{10, 50, 100\}$ have been plotted in the
        same figure.

@@

@@

>>>>

\title{Variational Monte Carlo on Bosonic Systems}
\author{Winther-Larsen, Sebastian Gregorius}
\homepage[Project code: ]{https://github.com/Schoyen/FYS4411}
\affiliation{University of Oslo}
\author{Schøyen, Øyvind Sigmundson}
\homepage[Project code: ]{https://github.com/Schoyen/FYS4411}
\affiliation{University of Oslo}
\date{\today}

\begin{abstract}
    In this project we use Quantum Variational Monte Carlo on a toy model of a  bosonic gas  in an elliptic trap. The framework
    we have built is first tested against analytic solutions of simple non-interacting systems. From there the system
    is extended to include perturbed harmonic oscillator traps and particle interaction. We compute one-body
    densities for the interacting case and discuss why this tool is important and how our toy model has similar 
    attributes to real-world systems.
\end{abstract}

@@

        The Laplacian of particle $k$ divided by the trial wavefunction is
        \begin{widetext}
            \begin{align}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                &=
                \frac{\nabla_k^2\phi_k}{\phi_k}
                + 2\frac{\nabla_k\phi_k}{\phi_k}
                \sum_{m \neq k}^N
                \frac{\Delta\vf{r}_{km}}{r_{km}}
                \dpd[]{u_{km}}{r_{km}}
                + \sum_{m\neq k}^N
                \para{
                    \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                    + \dpd[2]{u_{km}}{r_{km}}
                }
                +
                \para{
                    \sum_{m \neq k}^N
                    \frac{\Delta\vf{r}_{km}}{r_{km}}
                    \dpd[]{u_{km}}{r_{km}}
                }^2,
                \label{eq:laplacian_full_system}
            \end{align}
        \end{widetext}
        where $\Delta \vf{r}_{km} = \vf{r}_k - \vf{r}_m$ and $d$ is the
        dimensionality. Evaluating the gradient and Laplacian of the single
        particle functions yields
        \begin{gather}
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        		\Biggr)
       		 \Psi_T(\vf{r}).
        		\label{eq:gradient_full_wavefunction}
    		\end{gather}

@@

        Looking at the one-body part of the Hamiltonian we get
        \begin{align}
            H
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}\nabla_k^2
                + \half m\brak{
                    \omega^2(x_k^2 + y_k^2)
                    + \omega_z^2 z_k^2
                }}
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
             	    },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The single particle functions also get a scaling
            factor
            \begin{align}
                \phi_k
                &=
                \exp\brak{
                    -\frac{\alpha\hbar}{m\omega}
                    ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
                }
                \equiv
                \phi(\vf{r}_k').
                \label{eq:scaled_spf}
            \end{align}
            The correlation wavefunction and the interaction potential remain the
            same, but using the scaled hard core radius $a/a_{\text{ho}} =
            0.0043$.

@@

        \subsubsection{One-Body Densities}
            We now look at one-body densities for the value of $\alpha$ found in
            \autoref{eq:optimal_interacting_alpha}. The results for a system consisting
            of $n=100$ particles are displayed in \autoref{fig:one_body_density_100}.
            In this figure we see the densities for for a system confined to a 
            perturbed elliptic trap with and without interaction. We see that for the
            interacting case, the one-body density is skewed towards a larger 
            $\abs{\vf{r}}$, as is expected. The ``ideal'' density, a system with spherical
            trap, is also included in the figure.

@@

        That said, the way to get the best confirmation of the results presented 
        above is in comparison to experiment. This would be quite difficult
        as the systems presented are very theoretical in nature and most likely
        far from anything we would see in nature. Without being able to compare 
        our result with anything happening in the real world this entire study would
        have no value except for the instrumental value, being no more than
        a theoretical exercise.  However, the one-body
        density is a measure of the role of many-body correlations which in
        some cases can be related to experiment. For a system of charged 
        particles it can be used to determine a charge distribution, which in
        turn can be linked to experiment. It would be very fruitful to simulate a
        system of such particles that is more natural, i.e. real, but we realise the
        purpose of this study is first and foremost to introduce variational 
        Monte Carlo as a tool in computational physics. We assume, however,
        that this study is easy enough to expand into a study of alkali atoms like
        $^{87}Rb$, $^{23}Na$ or $^7Li$.

	\subsection{Energy per particle}
        For the non-interacting harmonic oscillator system there is no
        difference between a new ``particle'' and new dimensions
        \footnote{This is not entirely true in the case of our implementation of
        the Metropolis algorithms as we draw a single random particle and move
        all dimensions instead of iterating over all particles and all
        dimensions.}. This means that the inclusion of a new particle or
        dimension yields a linear scaling in the energy. Thus if we compute the
        energy per particle for systems of varying size we expect to get the
        same energy. This is most easily seen in \autoref{fig:initial_problem_b}
        and in the expression for the exact energy \autoref{eq:exact_energy}.
        This, however, is \emph{not} the case in the interacting system due to
        the Jastrow factor. Looking at \autoref{tab:10_interacting} and
        \autoref{tab:100_interacting} we see that multiplying the expected
        energy of the former with $10$ yields a lower number than the latter. We
        can also see this behavior in \autoref{fig:interacting}, where the
        energy per particle for $N = \{10, 50, 100\}$ have been plotted in the
        same figure.

@@

        The reason for the increase in energy per particle can be ascribed to the
        repulsive interaction added by the Jastrow factor. The repulsive force can
        be explain thusly; The Jastrow factor  increases the value of the 
        wavefunctions for any given particle if that particle is some distance away
         from any other particle. An increased value in the wavefucntion is equivalent
         to an increase in probability amplitude, again correpsonding to an increase in
         probability density, giving rise the the fact that particles are more likely to 
         appear further from one another.

	\subsection{Low density system}
		It is safe to say that the effect of adding a Jastrow factor is significant, 
		i.e. the increase in
        	energy for the system as the interaction is switched on is much higher
        	than the uncertainty of the energy. A simple look at \autoref{fig:interacting}
        	will be proof enough as the error bars of the figure are not discernible. 
        On the other hand, the energy added to the system by including interaction
        is not very high compared to the energy of the system as a whole. For the 
        larger of the interacting systems we have looked at, with $n=100$ particles,
        The interaction add only a mere $9\%$ extra energy (see \autoref{tab:10_optimal}).
        Studying \autoref{fig:one_body_density_100} will reveal the same.

        The reason for this stems from the assumption which justifies that
        adding only two-body interaction is sufficient to model all interaction to a
        reasonable extent. This assumption is justified by the fact that alkali and
        atomic hydrogen systems\footnote{Systems similar to our toy model}
        are dilute. For gases of alkali atoms confined in magnetic or optical traps
        the average distance between atoms is much larger than the range of
        inter-atomic interaction. The probability is high that atoms very seldom
        interact and if so, just with one other atom.

@@

	In this study we have built a functional variational Monte Carlo solver framework for 
	a gas following Bose-Einstein statistics confined in an elliptic trap. Monte Carlo 
	methods is a broad class of computational algorithms, whereas the variational
	Monte Carlo is a subset. Other applications of such methods within the physical
	sciences includes, but is not limited to, quantum chromodynamics, the modeling of
	aearodynamic forms, particle physics detector design and galaxy evolution. 
	Needless to say, Monte Carlo method is a valuable tool for a computational physicist.

	To prove the 
	validity of the framework we tested it against an analytically solved case, an ideal
	non-interacting system. Thereafter we have shown how our implementation can
	handle large systems of interacting particles and perturbed traps. Even though
	this study is somewhat theoretic in nature, we have discussed the possibility of
	expanding our machinery to a case bearing a larger resemblance to a system
	found in nature. To this end we have computed one-body densities for the 
	interacting system, which can be compared to real physical properties as
	measured in an experiment.
